---
title: "Week 6 Practical"
output: html_document
---

# Detecting Spatial Patterns

In this analysis we will analyse the patterns of Blue Plaques — you will see these placed on around the UK linking the buildings of the present to people of the past. - ***For any given London Borough, are the Blue Plaques within that borough distributed randomly or do they exhibit some kind of dispersed or clustered pattern?***

-   We want to find out are plaques at random? clustered? or dispersed?

```{r}
#clear memory
rm(list = ls()) 
gc()
```

```{r}
# load packages
library(spatstat)
library(here)
library(sp)
library(tmap)
library(sf)
library(tmaptools)
```

## Setting Up Data 

First, read in London Borough Boundaries (polygon)

-   We set the study area so that we can know where points occurs later and define "random" or "clustered"!

-   We also need this to clip Blue Plaques to London and tag each plaque with the borough

-   And other things...

```{r}
#London Borough Boundaries
LondonBoroughs <- st_read(here::here("Week 1 Practical Data", "statistical-gis-boundaries-london", "ESRI", "London_Borough_Excluding_MHW.shp"))

#or you can use this:
#LondonBoroughs <- st_read("https://opendata.arcgis.com/datasets/8edafbe3276d4b56aec60991cbddda50_4.geojson")
```

Second, we pull out London using `str_detect()` function from the `stringr` package in combination with `filter()` from `dplyr` (again!).

We will look for the bit of the district code that relates to London (E09) from the ‘lad15cd’ column data frame of our `sf` object

By doing this, we aim to achieve:

-   Filter study area (London - E09 only)

-   Project to meters: transform to 27700 (British National Grid so distances/ areas are in meters)

```{r}
library(stringr)
library(dplyr)

BoroughMap <- LondonBoroughs %>%
  dplyr::filter(str_detect(GSS_CODE, "^E09")) %>%
  st_transform(., 27700)

qtm(BoroughMap)
```

Ok, now lets give a quick sanity check before we start any spatial stats!

Good!! we see:

-   Row length =33, we got all 33 London Boroughs

-   Geometry type: MULTIPOLYGON: 33, all boroughs are multipolygons

-   Area distribution:

    -   min=3.15ha (That's tiny! should be City of London!), max=15,013ha(Bigger! maybe it's Bromley?)

    -   strong skew (mean 4,832\>median 358) (=mean is sensitive to big values)

```{r}
summary(BoroughMap)
```
